An Application of Graph Theory to MDAO Problem Formulation

### 2013-03-04 graph_syntax.tex
update for consolidation of node and edge types

We now assign additional properties to the nodes and edges so that they may represent the data flow of an MDAO problem. 
The first property assigned to nodes and edges is the \emph{type}. The possible node types are:
\begin{itemize}
\item global input
\item global output
\item local input
\item local output
\item analysis start
\item analysis finish
\end{itemize}
and the possible edge types are
\begin{itemize}
\item connection
\item in
\item analysis
\item out
\item ((consider eliminating `in' and `out' and using connection))
\end{itemize}
Then, considering sets $T_\txt{node}$ and $T_\txt{edge}$ to be sets containing the possible node types and edge types, respectively, we use mappings $t_\txt{node}:V \to T_\txt{node}$ and $t_\txt{edge}:E \to T_\txt{edge}$ to assign a type to each node and edge. Nodes and edges are named according to their type (e.g. connection edge). Additional properties will be discussed in subsequent sections.

Next, we introduce the graph $A$ which represents a single analysis block broken down into its individual inputs and outputs. An example graph is shown in Fig.~\ref{f:analysis block} to demonstrate the basic structure. This structure is derived from the need to represent each input and output individually while representing the actual analysis with a single node or edge. The `local input' nodes each represent a single input of the analysis block. Each of these nodes is connected to a single `analysis start' node via an `in' edge. Next the analysis start node is directed to the `analysis finish' node via the `analysis edge', which represents the actual analysis. Finally, the analysis finish node is directed to `local output' nodes via individual `out' edges.
\begin{figure}[htb!]
	\begin{center}
	\includegraphics[width=4in]{images/analysis_block}
	\end{center}
	\vspace{-20pt}
\caption{Example directed graph. The each node type and edge type is labeled, with nodes being label in italics.}
\label{f:analysis block}
\end{figure}
`Connection' edges connect local output nodes from one analysis block to either the local input nodes of other analysis blocks or global output node. In this way, the analysis block graph is the fundamental building block of the MDAO problem data flow.

\subsection{Maximal Connectivity Graph}
Now we define the node property \emph{variable name}, which represents the unique identity of the variables passed between analysis blocks as represented by nodes of type local input, local output, global input, and global output. Let $N$ be the set containing each unique variable name (no multiplicity), and let $n:V \to N$ be a mapping that assigns a variable name to a node.

Next, Consider an MDAO problem defined by a set of codes, global inputs, and global outputs, as represented by analysis blocks $A_i=(V_{A_i},E_{A_i}),i=1,\ldots,m$, the set of nodes $\mathcal I$, and the set of nodes $\mathcal O$, respectively. 
Then we may construct the \emph{maximal connectivity graph} $M=(V_M,E_M)$ as
\begin{IEEEeqnarray*}{rCl}
V_M & = & \mathcal I \cup \mathcal O \cup \left( \bigcup_{i = 1}^m V_{A_i} \right), \\
E_M & = & C_M \cup \left( \bigcup_{i=1}^m E_{A_i} \right),
\end{IEEEeqnarray*}
where $C$ is a set containing connection edges defined for $v_1,v_2 \in V_M$ as:
\begin{IEEEeqnarray*}{rl}
c = & (v_1,v_2) \in C_M \txt{ if:} \\
& t_\txt{node}(v_1) = \txt{`local output' or `global input'}, \\
& t_\txt{node}(v_2) = \txt{`local input' or `global output'},\txt{ and} \\
& n(v_1) = n(v_2).
\end{IEEEeqnarray*}
The maximal connectivity graph contains all possible connections that may exist between the given analysis blocks, and these connections are made by connecting variables with the same name, which requires a consistent naming convention between analysis blocks.

\subsection{Fundamental Problem Formulation Graph}
Before defining the the fundamental problem formulation graph, let us first define $\txt{deg}^-(v)$ as the \emph{indegree} $\txt{deg}^+(v)$ as the \emph{outdegree} of vertex $v$, where indegree and outdegree refer to the number of edges directed into vertex $v$ and the number of edges directed out of vertex $v$, respectively. Then, define a mapping $d:V \to \mathbb{N}$ to specifiy the maximum number of connections that are allowed into a local input node.
If $d(v)>1$ for a local input node, this means that multiple inputs are accepted, such as the case for a multifidelity formulation.

% Next, let $H$ be the set of all nodes which are \emph{holes} in $M$ and define $H$ as
% \begin{equation}
% H=\big\{v \in M \big| t_\txt{node}(v) = \txt{`local input' or `global output'} \txt{ and } \txt{deg}^-(v) = 0\big\}.
% \end{equation}
% Also, let $L$ be the set of all nodes which are \emph{collisions} in $M$ and define $L$ as
% \begin{equation}
% L = \big\{v \in M \big| t_\txt{node}(v) = \txt{`local input' or `global output'} \txt{ and } \txt{deg}^-(v) > d(v) \big\}.
% \end{equation}
We may now define the \emph{fundamental problem formulation graph}, $F=(V_F,E_F)$, as a directed graph meeting the following conditions
\begin{enumerate}
\item[(1)] $\displaystyle{V_F = \mathcal I_F \cup \mathcal O \cup \left( \bigcup_{i \in \mathcal A} V_{A_i} \right),\ \mathcal I_F \subset \mathcal I}$
\item[(2)] $\displaystyle{E_F = C_F \cup \left( \bigcup_{i \in \mathcal A} E_{A_i} \right)}$
\item[(3)] $\displaystyle{\forall v \in V_F \txt{ with } t_\txt{node}(v) = \txt{`local input' or `global output,'}\quad 0 < \txt{deg}^-(v) \leq d(v)}$
\end{enumerate}
The set $\mathcal A$ is an index set containing the indices of the analysis blocks in $F$. The first requirement for $F$ is that $V_F$ be composed of all of the global outputs (which are given as required), the nodes from each analysis block in $F$, and any global input nodes that are needed, $\mathcal I_F$. The second requirement suggests that the edges in $F$ comprise the edges for each analysis block and the connections between them, the global inputs, and the global outputs. The final requirement is specific to the number of connection edges directed into a local input node or a global outputs node. If there are no connections directed inward, the node is called a \emph{hole}, and if more connections are directed in than are allowed, the node is called a \emph{collision}.

\subsection{Obtaining the Fundamental Problem Formulation Graph}
In general, there may be multiple different graphs that satisfy the FPF conditions, though there may be none at all. Here, we describe a process for obtaining an FPF by starting with the MCG and removing nodes and edges until the FPF conditions are met. Then the problem is reduced to deciding which nodes and edges to remove, and this is centered on the local input nodes, which are treated as decision switches.
The first step is to detect holes and remove the corresponding analysis blocks, which could not be used in a

>>>>

We now assign additional properties to the nodes and edges so that they may represent the data flow of an MDAO problem. 
The first property assigned to nodes and edges is the \emph{type}. The possible node types are:
\begin{description}
\item[input] The `input' node has specific rules for the number of edges that may be directed into it.
\item[simple] The `simple' node has no additional properties and is used as a building block.
\end{description}

The possible edge types are:
\begin{description}
\item[connection] The `connection' edge represents the passing of data between analysis blocks.
\item[data] The `data' edge represents the passing of data within an analysis block.
\item[analysis] The `analysis' edge contains additional properties to represent the execution of an analysis block.
\end{description}
Then, considering sets $T_\txt{node}$ and $T_\txt{edge}$ to be sets containing the possible node types and edge types, respectively, we use mappings $t_\txt{node}:V \to T_\txt{node}$ and $t_\txt{edge}:E \to T_\txt{edge}$ to assign a type to each node and edge. Nodes and edges are herein referred to by their type (e.g. `connection' edge). Additional properties will be discussed in subsequent sections.

Next, we introduce the graph $A$ which represents a single analysis block broken down into its individual inputs and outputs. An example graph is shown in Fig.~\ref{f:analysis block} to demonstrate the basic structure. This structure is derived from the need to represent each input and output individually while representing the actual analysis with a single node or edge. The `input' nodes each represent a single input of the analysis block. Each of these nodes is connected to a single `simple' node via a `data' edge, which represents the gathering of the inputs to begin the analysis. Next the `simple' node is directed to another `simple' node via the `analysis edge', which represents the actual analysis. Finally, this `simple' node is directed to several `simple' nodes via individual `data' edges, and this represents individual local outputs which may then be connected to other analysis blocks.
\begin{figure}[htb!]
	\begin{center}
	\includegraphics[width=4in]{images/analysis_block}
	\end{center}
	\vspace{-20pt}
\caption{Example directed graph. The each node type and edge type is labeled, with nodes being label in italics.}
\label{f:analysis block}
\end{figure}
`Connection' edges connect the local outputs from one analysis block to input nodes. In this way, the analysis block graph is the fundamental building block of the MDAO problem data flow.

\subsection{Maximal Connectivity Graph}
Next, Consider an MDAO problem defined by a set of codes, global inputs, global outputs. The codes are represented by analysis blocks $A_i=(V_{A_i},E_{A_i}), \ i=1,\ldots,m$, the global inputs are represented a set of `simple' nodes $I$, and the global outputs are represented by a set of `input' nodes $O$. We assume that $O$, $I$, and $A_i$ are given, and that any potential connection edge is given in the form of the set $C_M$. 
Then we may construct the \emph{maximal connectivity graph} $M=(V_M,E_M)$ as
\begin{IEEEeqnarray*}{rCl}
V_M & = & I \cup O \cup \left( \bigcup_{i = 1}^m V_{A_i} \right), \\
E_M & = & C_M \cup \left( \bigcup_{i=1}^m E_{A_i} \right).
\end{IEEEeqnarray*}
The maximal connectivity graph contains all possible connections that may exist between the given analysis blocks.

\subsection{Fundamental Problem Formulation Graph}
Before defining the the fundamental problem formulation graph, let us first define $\txt{deg}^-(v)$ as the \emph{indegree} $\txt{deg}^+(v)$ as the \emph{outdegree} of vertex $v$, where indegree and outdegree refer to the number of edges directed into vertex $v$ and the number of edges directed out of vertex $v$, respectively. Then, define a mapping $d:V \to \mathbb{N}$ to specifiy the maximum number of connections that are allowed into an input node.
If $d(v)>1$ for an input node, this means that multiple inputs are accepted, such as the case for a multifidelity formulation.

% Next, let $H$ be the set of all nodes which are \emph{holes} in $M$ and define $H$ as
% \begin{equation}
% H=\big\{v \in M \big| t_\txt{node}(v) = \txt{`local input' or `global output'} \txt{ and } \txt{deg}^-(v) = 0\big\}.
% \end{equation}
% Also, let $L$ be the set of all nodes which are \emph{collisions} in $M$ and define $L$ as
% \begin{equation}
% L = \big\{v \in M \big| t_\txt{node}(v) = \txt{`local input' or `global output'} \txt{ and } \txt{deg}^-(v) > d(v) \big\}.
% \end{equation}
We may now define the \emph{fundamental problem formulation graph}, $F=(V_F,E_F)$, as a directed graph meeting the following conditions
\begin{enumerate}
\item[(1)] $\displaystyle{V_F = I_F \cup O \cup \left( \bigcup_{i \in \mathcal A} V_{A_i} \right),\ I_F \subset I}$
\item[(2)] $\displaystyle{E_F = C_F \cup \left( \bigcup_{i \in \mathcal A} E_{A_i} \right)}$
\item[(3)] $\displaystyle{\forall v \in V_F \txt{ with } t_\txt{node}(v) = \txt{`input,'}\quad 0 < \txt{deg}^-(v) \leq d(v)}$
\end{enumerate}
The set $\mathcal A$ is an index set containing the indices of the analysis blocks in $F$. The first requirement for $F$ is that $V_F$ be composed of all of the global outputs (which are given as required), the nodes from each analysis block in $F$, and any global input nodes that are needed, $I_F$. The second requirement suggests that the edges in $F$ comprise the edges for each analysis block and the connections between them, the global inputs, and the global outputs. The final requirement is specific to the number of connection edges directed into a local input node or a global outputs node. If there are no connections directed inward, the node is called a \emph{hole}, and if more connections are directed in than are allowed, the node is called a \emph{collision}. The final requirement is therefore a requirement on $C_F$.

\subsection{Obtaining the Fundamental Problem Formulation Graph}
In general, there may be multiple different graphs that satisfy the FPF conditions, though there may be none at all. Here, we describe a process for obtaining an FPF by starting with the MCG and removing nodes and edges until the FPF conditions are met. Then the problem is reduced to deciding which nodes and edges to remove, and this is centered on the input nodes, which are treated as decision switches.

Then, mathematically, the FPF starts with $\mathcal A_1 = \{1,\ldots,m\}$, $I_{F,1} = I$, and $C_{F,1} = C_M$.

\begin{enumerate}
\item The first step is to detect holes and remove the corresponding analysis blocks, which could not be used because the required inputs are not supplied.
\begin{equation}
\forall i \in \mathcal A_1,\txt{ and }\forall v \in V_{A_i},\txt{ if deg}^-(v)=0 \txt{ then } i \in \mathcal B
\end{equation}
\begin{equation}
\mathcal A_2 = \mathcal A_1 \setminus \mathcal B
\end{equation}
\end{enumerate}
